---
title: "Seq_analysis"
author: "Alice Osmaston"
date: "2024-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

```{r}
# Load necessary packages
pacman::p_load(tidyverse, dplyr, knitr, here, ggplot2)

# Import the datasets from pore occupancy plots and pore scan plots
poreoccupancy <- read_csv(here("data", "Pore_activity_%_pores_sequencing.csv"))
porecount <- read_csv(here("data", "porecount2.csv"))
```


```{r}
# Plot the raw data
ggplot(data=poreoccupancy, aes(x=Hour, y=`% of pores sequencing (pore activity)`, color=Group))+ geom_point() + labs(title= "Percentage of sequencing pores over time (from pore occupancy plot)") + theme_minimal()
```



```{r}
# plot the data using a log transformation on the y axis to make the data linear
ggplot(data=poreoccupancy, aes(x=Hour, y=log(`% of pores sequencing (pore activity)`),color=Group))+ geom_point() + labs(title= "Percentage of sequencing pores over time (from pore occupancy plot)") + theme_minimal()
```
```{r}
# simple linear regression model
linear_model_occupancy <- lm(log(`% of pores sequencing (pore activity)`) ~ Hour * Group, data = poreoccupancy)

# summarise simple linear regression model
summary(linear_model_occupancy)

# Plot overlaid linear model
ggplot(data=poreoccupancy, aes(x=Hour, y=log(`% of pores sequencing (pore activity)`),color=Group))+ geom_point() + labs(title= "Percentage of sequencing pores over time (from pore occupancy plot)") + geom_smooth(method = "lm", formula = y ~ x, se = FALSE) + theme_minimal()
```


Baseline = -0.077025 which is (log) rate of decline in the reference group (phenol fragmented).
Add adjusted rates to the baseline to calculate the slowest rate of decline:

- **Phenol Fragmented:**
  = `-0.077025`
- **Phenol No Fragmented:**  
  `-0.017183 + -0.077025` = `r -0.017183 + -0.077025`
- **Zymo Kit Fragmented:**  
  `0.037087 + -0.077025` = `r 0.037087 + -0.077025`
- **Zymo Kit No Fragmented:**  
  `0.004085 + -0.077025` = `r 0.004085 + -0.077025`


```{r}
# Fit a GLM with a gamma family
gamma_glm_model_occupancy <- glm(`% of pores sequencing (pore activity)` ~ Hour * Group,
                 family = Gamma(link = "log"), data = poreoccupancy)

# Check the summary of the model
summary(gamma_glm_model_occupancy)

# Plot the GML with gamma link over the plot
ggplot(data = poreoccupancy, aes(x = Hour, y = log(`% of pores sequencing (pore activity)`), color = Group)) +
  geom_point() +
  labs(title = "Percentage of sequencing pores over time (from pore occupancy plot)") +
  stat_smooth(method = "glm", method.args = list(family = Gamma(link = "log")), formula = y ~ x, se = FALSE) +
  theme_minimal()

```
Baseline = -0.0766737 which is (log) rate of decline in the reference group (phenol fragmented).
Add adjusted rates to the baseline to calculate the slowest rate of decline:

- **Phenol Fragmented:**
  = `-0.0766737`
- **Phenol No Fragmented:**  
  `-0.0167150 + -0.0766737` = `r -0.0167150 + -0.0766737`
- **Zymo Kit Fragmented:**  
  `0.0369964 + -0.0766737` = `r 0.0369964 + -0.0766737`
- **Zymo Kit No Fragmented:**  
  `0.0047661 + -0.0766737` = `r 0.0047661 + -0.0766737`



```{r}

```

