---
title: "Seq_analysis"
author: "Alice Osmaston"
date: "2024-04-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

```{r}
# Load necessary packages
pacman::p_load(tidyverse, dplyr, knitr, here, ggplot2)

# Import the datasets from pore occupancy plots and pore scan plots
poreoccupancy <- read_csv(here("data", "Pore_activity_%_pores_sequencing.csv"))
porecount2 <- read_csv(here("data", "porecount2.csv"))
```

# Plot the pore count data (number of active pores)

```{r}
# Plot the raw data
ggplot(data=porecount2, aes(x=Hour, y=`Number of available pores`, color=Group))+ geom_point() + labs(title= "Number of active pores over time") + theme_minimal()
```

The data was log transformed to linearise it.

```{r}
# plot the data using a log transformation on the y axis to make the data linear
ggplot(data=porecount2, aes(x=Hour, y=log(`Number of available pores`),color=Group))+ geom_point() + labs(title= "Number of active pores over time") + theme_minimal()
```

Edit data to crop the first hour (hour 0) from each run. Because in some cases the number of available rises before falling.
Later we will also removed the final part of the run, where percentage
of pores sequencing is variable and non-informative as it tails off
towards zero.


```{r}
# Import dataset with first hour removed and normalised pore count column adjusted to the value in hour=1. 
porecount3 <- read_csv(here("data", "porecount3.csv"))
```


```{r}
# Apply simple linear regression model
lm_1 <- lm(log(`Number of available pores`) ~ Hour * Group, data = porecount3)

# summarise simple linear regression model
summary(lm_1)

# Plot overlaid linear model
ggplot(data=porecount3, aes(x=Hour, y=log(`Number of available pores`),color=Group))+ geom_point() + labs(title= "Number of available pores over time") + geom_smooth(method = "lm", formula = y ~ x, se = FALSE) + theme_minimal()
```
Interpretation of linear model coefficients:
Rate of decrease of available pores in Phenol Fragmented group = Baseline = -0.074493 
Phenol No Fragmented = -0.074493 + -0.006560 = `r -0.074493 + -0.006560`
Zymo Kit Fragmented = -0.074493 + 0.031813 = `r -0.074493 + 0.031813`
Zymo Kit No Fragmented = -0.074493 + 0.015106 = `r -0.074493 + 0.015106`

## Phenol No Fragmented vs Zymo No Fragmented. RESULT = Zymo No Fragmented is better (sig = ***)

```{r}
lm_comparison1 <- lm(log(`Normalised number of pores`) ~ Hour * Group,
                              data = porecount3[porecount3$Group == 'Zymo Kit No Fragmented' | 
                                                 porecount3$Group == 'Phenol No Fragmented',])
summary(lm_comparison1)
```

## Phenol No Fragmented vs Phenol Fragmented. RESULT = Phenol fragmented is better (sig = *)

```{r}
lm_comparison2 <- lm(log(`Normalised number of pores`) ~ Hour * Group,
                              data = porecount3[porecount3$Group == 'Phenol Fragmented' | 
                                                 porecount3$Group == 'Phenol No Fragmented',])
summary(lm_comparison2)
```


Note - the following hours are less than the max_hours calculated above
because each pore scan is done ever 1hr4 mins so there are less
datapoints than for the occupancy data. r 'print(max_hours)\`


# Remove final seqeuncing hours from each condition after 90% of data has been generated

Now cut off the final sequencing hours which is uninformative and highly
variable and disperse and only accounts for a small amount of the data
output. We keep only the linear region of each plot.

Number of hours that each run lasted:

r `print(max_hours)`

Now cut off the final part of the sequencing run to remove variablitiy
and keep only the linear region of each plot.

Hours to cut off from each run:

-   **Phenol Fragmented:**\
    90% = Cut off 10 hours
-   **Phenol No Fragmented:**\
    90% = Cut off 18 hours
-   **Zymo Kit Fragmented:**\
    90% = Cut off 27 hours
-   **Zymo Kit No Fragmented:**\
    90% = Cut off 17 hours

```{r}
# Calculate the maximum hour for each group
max_hours <- porecount3 %>%
  group_by(Group) %>%
  summarise(MaxHour = max(Hour))
# Merge the maximum hours with the original data
porecount3 <- porecount3 %>%
  left_join(max_hours, by = "Group")
```

```{r}
# Applying the crop conditions for each group
cropped_90_count <- porecount3 %>%
  filter(
    (Group == "Zymo Kit Fragmented" & Hour <= (MaxHour - 27)) |
    (Group == "Zymo Kit No Fragmented" & Hour <= (MaxHour - 17)) |
    (Group == "Phenol No Fragmented" & Hour <= (MaxHour - 18)) |
    (Group == "Phenol Fragmented" & Hour <= (MaxHour - 10))
  )

```


```{r}
# simple linear regression model
lm2 <- lm(log(`Normalised number of pores`) ~ Hour * Group, data = cropped_90_count)

# summarise simple linear regression model
summary(lm2)

# Plot overlaid linear model
ggplot(data=cropped_90_count, aes(x=Hour, y=log(`Normalised number of pores`),color=Group))+ geom_point() + labs(title= "Normalised count of sequencing pores over time (90% of data)") + geom_smooth(method = "lm", formula = y ~ x, se = FALSE) + theme_minimal()
```
Interpretation of linear model coefficients:
Rate of decrease of available pores in Phenol Fragmented group = Baseline = -0.086804 
Phenol No Fragmented = -0.086804 + -0.010362 = `r -0.086804 + -0.010362`
Zymo Kit Fragmented = -0.086804 + 0.043981 = `r -0.086804 + 0.043981`
Zymo Kit No Fragmented = -0.086804 + 0.009686 = `r -0.086804 + 0.009686`



## Phenol No Fragmented vs Zymo No Fragmented. RESULT = Zymo No Fragmented is better (sig = ***)

```{r}
lm_comparison3 <- lm(log(`Normalised number of pores`) ~ Hour * Group,
                              data = cropped_90_count[cropped_90_count$Group == 'Zymo Kit No Fragmented' | 
                                                 cropped_90_count$Group == 'Phenol No Fragmented',])
summary(lm_comparison3)
```

## Phenol No Fragmented vs Phenol Fragmented. RESULT = Phenol fragmented is better (sig = ***)

```{r}
lm_comparison4 <- lm(log(`Normalised number of pores`) ~ Hour * Group,
                              data = cropped_90_count[cropped_90_count$Group == 'Phenol Fragmented' | 
                                                 cropped_90_count$Group == 'Phenol No Fragmented',])
summary(lm_comparison4)
```